---
title: "Introduction, Chapter 1 and Chapter 2"
author: "Joe Shaw"
format: html
editor: visual
---

"The details of R's memory management are not documented in a single place. ... I figured out by reading the C source code".

So some of R is written in C. Wikipedia: "It is written primarily in C, Fortran, and R itself."

## Timeline

https://en.wikipedia.org/wiki/S\_%28programming_language%29

https://en.wikipedia.org/wiki/R\_(programming_language)

-   1976: S was a language built at Bell Laboratories (New Jersey, USA) - the S stood for "Statistics".

-   1980: S distributed outside of Bell Laboratories.

-   1988: the "new S language" S-PLUS released.

-   1993: Ross Ihaka and Robert Gentleman develop R, based on S, to teach statistics at the University of Auckland (New Zealand).

-   1997: R becomes a GNU project - free, publicly accessible and built by mass collaboration.

-   2000: first official version (R v1.0) released.

So initially the language was devloped by a private company (Bell Labs) then modified at a public university, and then made freely available by the public-access GNU project.

## Assignment

```{r}
#| include: TRUE

x <- c(1, 2, 3)

```

Inaccurate translation into English: "create an object named 'x', containing the values 1, 2, and 3".

More accurate translation into English: "create an object (vector of values with 1, 2 and 3), then **bind** that object to a name, which is x".

The value doesn't have a name. The name has a value.

So (annoyingly) the direction of binding is the opposite of the assignment operator "name \<- value". Because actually you're saying "this name has the following value", which would be easier to write as "name -\> value".

*"The arrow points in opposite direction to the assignment arrow: \<- creates a binding from the name on the left-hand side to the object on the right-hand side."*

"Thus, you can think of a name as a reference to a value. For example, if you run this code, you don't get another copy of the value c(1, 2, 3), you get another binding to the existing object:"

```{r}
#| include: TRUE

x <- c(1, 2, 3)

y <- x

```

There is still only one "version" of c(1, 2, 3) but now 2 names are being pointed at it.

```{r}

library(lobstr)

lobstr::obj_addr(mean)
lobstr::obj_addr(base::mean)
lobstr::obj_addr(get("mean"))
lobstr::obj_addr(evalq(mean))
lobstr::obj_addr(match.fun("mean"))

```

## Syntactic Names

Syntactic name: a name which includes characters that are syntactically valid - i.e. do not contain special characters, protected words like TRUE, or begin with a number.

ASCII: American Standard Code for Information Interchange - a standard of the different characters that can be used in computing. Wikipedia: first edition published in 1963, most recently updated in 1986.

There are 128 ASCII characters, and only 96 of those are the ones we commonly use.

That means that the all of computer programming is the arrangement of 96 characters in different sequences.

! " \# \$ % & ' ( ) \* + , - . / 0 1 2 3 4 5 6 7 8 9 : ; \< = \> ? \@ A B C D E F G H I J K L M N O P Q R S T U V W X Y Z \[   \] \^ \_ \` a b c d e f g h i j k l m n o p q r s t u v w x y z { \| } \~

-   Lowercase alphabet letters 26

-   Upper case alphabet letters 26

-   Punctuation (!, space, :, ; , comma, full stop, &, ?, \`) 9

-   Mathematical operators (+, \*, -, /, \^, =, %, \<, \>, \|) 10

-   Brackets (two normal, two square, two curly) 6

-   Other (#, \$, \_, \~) 4

-   Backslash and forward slash 2

## Copy on modify

Copy on modify: best illustrated by the following code

```{r}
#| include: TRUE

x <- c(1, 2, 3)

y <- x
# x and y both are both bound to the same vector. We can tell this 
# by checking the object address.

lobstr::obj_addr(x) == lobstr::obj_addr(y)

# Then we modify y
y[[3]] <- 4

# Because we've modified y, a new version of the vector is made 
# (copied) so that x is not changed as well. In this way, the 
# original vector is "immutable" - it cannot be changed without copying it.

lobstr::obj_addr(x) == lobstr::obj_addr(y)

print(x)
print(y)

```

"Like vectors, lists use copy-on-modify behaviour; the original list is left unchanged, and R creates a modified copy. This, however, is a shallow copy: the list object and its bindings are copied, but the values pointed to by the bindings are not."

"Data frames are lists of vectors"

Copy on modify is good for reducing the amount of computer memory required for computations.

## Object Size

"You can find out how much memory an object takes with lobstr::obj_size()"

"Because R uses a **global string pool** character vectors take up less memory than you might expect: repeating a string 100 times does not make it take up 100 times as much memory."

```{r}
#| include: TRUE

banana <- "bananas bananas bananas"

obj_size(banana)

obj_size(rep(banana, 100))

```

## Modify in Place

If there's only one name bound to an object, and the object is modified, then R doesn't perform copy on modify. Instead it just modifies the object, which is called "modify in place".

## Garbage Collector

The garbage collector "frees up memory by deleting R objects that are no longer used, and by requesting more memory from the operating system if needed."

"The garbage collector (GC) runs automatically whenever R needs more memory to create a new object. Looking from the outside, it's basically impossible to predict when the GC will run. In fact, you shouldn't even try."

## Key point

There are lots of diagrams included in the textbook for this chapter that are really helpful.
