---
title: "Chapter 4: Subsetting"
author: "Joe Shaw"
format: html
---

"Subsetting in R is easy to learn but hard to master because you need to internalise a number of interrelated concepts:

- There are six ways to subset atomic vectors.

- There are three subsetting operators, [[, [, and $.

- Subsetting operators interact differently with different vector types (e.g., atomic vectors, lists, factors, matrices, and data frames).

- Subsetting can be combined with assignment."

# Selecting Multiple Elements

You can supply number vectors within [] subsetting to select multiple atomic vector elements.

**Positive integers return elements at the specified positions**

```{r}
#| include: TRUE

x <- c(2.1, 4.2, 3.3, 5.4)

# Select positions 3 and 1 from x
x[c(3, 1)]

# Select poisiton 1 twice
x[c(1, 1)]

# Real numbers are silently truncated to integers
# So selecting position 2.1 and 2.9 returns position 2 twice
# (there is no rounding up)
x[c(2.1, 2.9)]

```

**Negative integers exclude elements at the specified positions**

```{r}
#| include: TRUE
#| error: TRUE

# Excludes positions 3 and 1
x[-c(3, 1)]

# Same result
x[c(-3, -1)]

# You can't mix positive and negative integers
x[c(-1, 2)]


```

**Logical vectors select elements where the corresponding logical value is TRUE**

```{r}
#| include: TRUE

# Selects the first two elements, but not the third and fourth
x[c(TRUE, TRUE, FALSE, FALSE)]

# Select elements which have a value above 3
x[x > 3]

```

**Nothing returns the original vector. **

```{r}
#| include: TRUE

x[]

```

**Zero returns a zero-length vector.**

```{r}
#| include: TRUE

x[0]

```

# Matrices and Arrays

You can subset lists by supplying a vector for each dimension.

```{r}
#| include: TRUE

a <- matrix(1:9, nrow = 3)

colnames(a) <- c("A", "B", "C")

a

# Select rows 1 and 2, and all columns
a[1:2, ]

# Select rows 1 and 3, and columns B and A
a[c(TRUE, FALSE, TRUE), c("B", "A")]

# Select 0 rows and columns A and C
a[0, -2]

```

"Arrays in R are stored in column-major order"

```{r}
#| include: TRUE

vals <- outer(1:5, 1:5, FUN = "paste", sep = ",")

vals

# Select element 4 and 15, which are identified by counting downwards from position
# 1, 1
vals[c(4, 15)]

```

# Data frames and tibbles

When subsetting with a single index, data frames behave like lists.

When subsetting with two indices, they behave like matrices.

```{r}
#| include: TRUE

df <- data.frame(x = 1:3, y = 3:1, z = letters[1:3])

df[df$x == 2, ]

# Select rows 1 and 3
df[c(1, 3), ]

```

"By default, subsetting a matrix or data frame with a single number, a single name, or a logical vector containing a single TRUE, will simplify the returned output, i.e. **it will return an object with lower dimensionality**. To preserve the original dimensionality, you must use drop = FALSE."

"Data frames with a single column will return just the content of that column"

```{r}
#| include: TRUE

df <- data.frame(a = 1:2, b = 1:2)

str(df[, "a"])


str(df[, "a", drop = FALSE])

```

"The default drop = TRUE behaviour is a common source of bugs in functions: you check your code with a data frame or matrix with multiple columns, and it works. Six months later, you (or someone else) uses it with a single column data frame and it fails with a mystifying error."



```{r}
#| include: TRUE

# Exercises

library(tidyverse)

mtcars[mtcars$cyl == 4, ]

mtcars[-c(1:4), ]

mtcars[mtcars$cyl <= 5,]

mtcars[mtcars$cyl %in% c(4, 6), ]

```

# Selecting a single element

"[[ is used for extracting single items, while x$y is a useful shorthand for x[["y"]]."

```{r}
#| include: TRUE
#| error: TRUE

var <- "cyl"

# Doesn't work - mtcars$var translated to mtcars[["var"]]
mtcars$var

# Instead use [[
mtcars[[var]]

```

# Subassignment

"All subsetting operators can be combined with assignment to modify selected values of an input vector: this is called subassignment."

Logical subsetting

```{r}
#| include: TRUE

mtcars[mtcars$gear == 5, ]

mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]

```

De Morgan’s laws:

- !(X & Y) is the same as !X | !Y
- !(X | Y) is the same as !X & !Y
