---
title: "Data Tidying"
author: "Joe Shaw"
date: "26/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

*"There are three interrelated rules that make a dataset tidy:*

*- Each variable is a column; each column is a variable.*

*- Each observation is a row; each row is an observation.*

*- Each value is a cell; each cell is a single value."*

## Lengthening Data

There are two common reasons why data is untidy.

*"Data is often organized to facilitate some goal other than analysis. For example, it’s common for data to be structured to make data entry, not analysis, easy."*

*"Most people aren’t familiar with the principles of tidy data, and it’s hard to derive them yourself unless you spend a lot of time working with data."*

```{r, include = TRUE}

head(billboard)

colnames(billboard)

billboard_longer <- billboard |> 
  pivot_longer(
    # Specifies the original columns which need to be pivoted. I.e. the columns that aren't variables, but instead are values
    # Uses same syntax as select
    cols = starts_with("wk"), 
    # What we want to call the variable of the values that are stored in the original column names
    names_to = "week", 
    # What we want to call the variable of the values that are stored in the original old columns
    values_to = "rank",
    #"These NAs don’t really represent unknown observations; they were forced to exist by the structure of the dataset"
    values_drop_na = TRUE
  )

billboard_longer

```

```{r, include = TRUE}

# Use parse_number() to extract numbers from strings.

billboard_longer |> 
  mutate(week = parse_number(week))

# Plot results
billboard_longer |> 
  mutate(week = parse_number(week)) |> 
  ggplot(aes(x = week, y = rank, group = track)) +
  geom_line(alpha = 0.25) +
  scale_y_reverse()

```


```{r, include = TRUE}

# tribble() can be used to make small dataframes row by row

df <- tribble(
  ~id,  ~bp1, ~bp2,
   "A",  100,  120,
   "B",  140,  115,
   "C",  120,  125
)

df1 <- data.frame(
  
  "id" =  c("A", "B", "C"),
  "bp1" = c(100, 140, 120),
  "bp2" = c(120, 115, 125)

)

all_equal(df1, df)

df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )

```

```{r, include = TRUE}

head(who2)

colnames(who2)

who2 |> 
  pivot_longer(
    # Pivot all the columns except country and year
    cols = !(country:year),
    # New column names
    names_to = c("diagnosis", "gender", "age"), 
    # Separate the values in the old column names by the underscore
    names_sep = "_",
    # Values in old columns become the "count" column
    values_to = "count"
  )

```

```{r, include = TRUE}

# This dataframe has child, dob and name all stored in column names, separated by underscores
household

colnames(household)

# These column names have both variable names (dob, name) and values (child1, child2) stored in them.

# So we need a way to separate them.

household |> 
  pivot_longer(
    cols = !family, 
    # The bits of the column to separate are separated by an underscore
    names_sep = "_", 
    # The ".value" bit is important
    # The column name structure is the first bit which is a variable (dob or name), then an underscore, then a bit which we'll call "child".
    names_to = c(".value", "child"), 
    # There is no "values_to" argument
    values_drop_na = TRUE
  )

```

## Using pivot_wider

```{r, include = TRUE}

str(cms_patient_experience)

cms_patient_experience |> 
  distinct(measure_cd, measure_title)

```

*"pivot_wider() has the opposite interface to pivot_longer(): instead of choosing new column names, we need to provide the existing columns that define the values (values_from) and the column name (names_from)"*

pivot_longer = names_to    values_to    cols
pivot_wider =  names_from  values_from  id_cols

```{r, include = TRUE}

cms_patient_experience |> 
  pivot_wider(
    # Specify which columns have values that uniquely identify each row
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )

```

## How pivot_wider works

pivot_wider performs the following steps:

1. Gets the new column names from the unique values in the original column supplied to names_from argument of pivot_wider ("measurement" in the below example)
2. Gets the new rows from the unique values in the original column(s) supplied to id_cols argument of pivot_wider ("id" in the below example)
3. The new column names and new rows are combined to generate an empty data frame with NA values
4. The missing values are then filled in using the values in the original column supplied to the values_from argument of pivot_wider ("number" in the example below).


```{r, include = TRUE}

df <- tribble(
  ~id, ~measurement, ~number,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)

df |> 
  pivot_wider(
    # Make new columns named bp1, bp2 etc
    names_from = measurement,
    values_from = number
  )

```

The strange thing is that pivot_wider can therefore "create" missing values by inserting NAs into a dataset. For example, there is no third blood pressure value (bp3) for id B in the original dtatframe in the example above. But the structure of the wider version forces an NA to be inserted, making this more obvious.

```{r, include = TRUE}

way1 <- df |> 
  distinct(measurement) |> 
  # "pull" extracts a single column
  pull()

way2 <- unique(df$measurement)

way1 == way2

# Both ways give the same result, but distinct is from dplyr and therefore works with pipes, whereas unique is from base R.

```

*"Another challenge is that, for a given dataset, it can be impossible to label the longer or the wider version as the “tidy” one. This is partly a reflection of our definition of tidy data, where we said tidy data has one variable in each column, but we didn’t actually define what a variable is (and it’s surprisingly hard to do so). It’s totally fine to be pragmatic and to say a variable is whatever makes your analysis easiest. So if you’re stuck figuring out how to do some computation, consider switching up the organisation of your data; don’t be afraid to untidy, transform, and re-tidy as needed!"*

## Diagrams

Diagrams are saved on Dropbox int his folder:

Dropbox/Documents/R projects/r_for_data_science_2e_images
