---
title: "Scripts and Projects"
author: "Joe Shaw"
date: "26/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(readxl)

```

Shortcut: Ctrl+Shift+S

This runs your entire script.

*"You should never include install.packages() in a script you share. It’s inconsiderate to hand off a script that will change something on their computer if they’re not being careful!"*

## Filenames

*"Three important principles for file naming are as follows:"*

*-"File names should be machine readable: avoid spaces, symbols, and special characters. Don’t rely on case sensitivity to distinguish files."*

*-"File names should be human readable: use file names to describe what’s in the file.*

*-"File names should play well with default ordering: start file names with numbers so that alphabetical sorting puts them in the order they get used."*

## Projects

# What is the source of truth?

Your source of truth should be R scripts, not your environment.

R scripts and data files can recreate the environment.

*"we highly recommend that you instruct RStudio not to preserve your workspace between sessions."*

Shortcut: Ctrl + Shift + F10 to restart R

To make sure you are capturing truth in the editor, not the environment, you can restart R and then rerun the script.

For this, use the shortcuts Ctrl + Shift + F10 then Ctrl + Shift + S.

*"We collectively use this pattern hundreds of times a week."*

# Where does your analysis live?

*"You can set the working directory from within R but **we do not recommend it**:"*

Better to set the working directory within a project file..

Windows uses backslash within file paths, which is annoying because in R the backslash is an escape character.

```{r, include = TRUE}

# Windows filepath
# "C:\Users\joseph.shaw2\Documents\r_for_data_science_2e\data\illumina_flow_cells.xlsx"

# Windows filepath with backslash escapes
path1 <- "C:\\Users\\joseph.shaw2\\Documents\\r_for_data_science_2e\\data\\illumina_flow_cells.xlsx"

read_excel(path = path1)

# Windows filepath with backslashes changed to forward slashes
path2 <- "C:/Users/joseph.shaw2/Documents/r_for_data_science_2e/data/illumina_flow_cells.xlsx"

read_excel(path2)

```