---
title: "Layers"
author: "Joe Shaw"
date: "15/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

*"Mapping an unordered discrete (categorical) variable (class) to an ordered aesthetic (size or alpha) is generally not a good idea because it implies a ranking that does not in fact exist."*

*"R has 25 built-in shapes that are identified by numbers. There are some seeming duplicates: for example, 0, 15, and 22 are all squares. The difference comes from the interaction of the color and fill aesthetics. The hollow shapes (0–14) have a border determined by color; the solid shapes (15–20) are filled with color; the filled shapes (21–24) have a border of color and are filled with fill. Shapes are arranged to keep similar shapes next to each other."*

*"Every geom function in ggplot2 takes a mapping argument, either defined locally in the geom layer or globally in the ggplot() layer. However, not every aesthetic works with every geom."*

*"ggplot2 will automatically group the data for these geoms whenever you map an aesthetic to a discrete variable "*

*"If you place mappings in a geom function, ggplot2 will treat them as local mappings for the layer. It will use these mappings to extend or overwrite the global mappings for that layer only. This makes it possible to display different aesthetics in different layers."*

*"ggplot2 provides more than 40 geoms but these don’t cover all possible plots one could make. If you need a different geom, we recommend looking into extension packages first to see if someone else has already implemented it"*

The graphical argument used to specify point shapes is pch.

## Facets

You can facet by multiple variables!

*"To facet your plot with the combination of two variables, switch from facet_wrap() to facet_grid()."*

```{r, include = TRUE}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv ~ cyl)

```

facet_grid() follows the format of facet_grid(rows ~ columns).

Using . tells it not to facet by either rows or columns.

```{r, include = TRUE}

# Don't facet by columns
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

# Don't facet by rows
ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)

```

## Statistical Transformations

*"Bar charts, histograms, and frequency polygons bin your data and then plot bin counts, the number of points that fall in each bin."*

*"Smoothers fit a model to your data and then plot predictions from the model."*

*"Boxplots compute the five-number summary of the distribution and then display that summary as a specially formatted box."*

*"Every geom has a default stat; and every stat has a default geom."*

*"To find the possible variables that can be computed by the stat, look for the section titled “computed variables” in the help for geom_bar()."*

You can use position = jitter with geom_point to avoid the problem of multiple points overlying eachother.

```{r, include = TRUE}

ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(position = "jitter")

```

*"Adding randomness seems like a strange way to improve your plot, but while it makes your graph less accurate at small scales, it makes your graph more revealing at large scales. "*

## Coordinate Systems

*"coord_quickmap() sets the aspect ratio correctly for geographic maps."*

## Grammar of Graphics

*"We can expand on the graphing template you learned in Section 1.3 by adding position adjustments, stats, coordinate systems, and faceting"*

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <COORDINATE_FUNCTION> +
  <FACET_FUNCTION>

*"In practice, you rarely need to supply all seven parameters to make a graph because ggplot2 will provide useful defaults for everything except the data, the mappings, and the geom function."*

*"The grammar of graphics is based on the insight that you can uniquely describe any plot as a combination of a dataset, a geom, a set of mappings, a stat, a position adjustment, a coordinate system, a faceting scheme, and a theme."*
