---
title: "Communication"
author: "Joe Shaw"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)

```

*"The purpose of a plot title is to summarize the main finding. Avoid titles that just describe what the plot is, e.g., “A scatterplot of engine displacement vs. fuel economy”."*

*"It’s possible to use mathematical equations instead of text strings. Just switch "" out for quote() and read about the available options in ?plotmath:"*

```{r, include = TRUE}

df <- tibble(
  x = 1:10,
  y = cumsum(x^2)
)

ggplot(df, aes(x, y)) +
  geom_point() +
  labs(
    x = quote(x[i]),
    y = quote(sum(x[i] ^ 2, i == 1, n))
  )

```

## Annotations

*"geom_text() is similar to geom_point(), but it has an additional aesthetic: label. This makes it possible to add textual labels to your plots."*

*"There are two possible sources of labels. First, you might have a tibble that provides labels."* The second option is to use annotate.

*"Another handy function for adding annotations to plots is annotate(). As a rule of thumb, geoms are generally useful for highlighting a subset of the data while annotate() is useful for adding one or few annotation elements to a plot."*

```{r, include = TRUE}

label_info <- mpg |>
  group_by(drv) |>
  arrange(desc(displ)) |>
  slice_head(n = 1) |>
  mutate(
    drive_type = case_when(
      drv == "f" ~ "front-wheel drive",
      drv == "r" ~ "rear-wheel drive",
      drv == "4" ~ "4-wheel drive"
    )
  ) |>
  select(displ, hwy, drv, drive_type)

ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.3) +
  geom_smooth(se = FALSE) +
  geom_label_repel(
    data = label_info, 
    aes(x = displ, y = hwy, label = drive_type),
    fontface = "bold", size = 5, nudge_y = 2
  ) +
  theme(legend.position = "none")

```

## Scales

*"Normally, ggplot2 automatically adds scales for you."*

*"Collectively axes and legends are called guides. Axes are used for x and y aesthetics; legends are used for everything else."*

Scales have the name format, scale plus axis plus scale type:

scale_x_continuous

scale_x_log10

## Legends

You can make the points in a legend bigger using the override.aes argument in guides.

```{r, include = TRUE}

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2, override.aes = list(size = 4)))

```

You can specify specific colours for specific values, rather than having to reorder factor levels.

```{r, include = TRUE}

presidential |>
  mutate(id = 33 + row_number()) |>
  ggplot(aes(x = start, y = id, color = party)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_color_manual(values = c(Republican = "#E81B23", Democratic = "#00AEF3"))

```

## Zooming

*"There are three ways to control the plot limits:*

*-Adjusting what data are plotted.*

*-Setting the limits in each scale.*

*-Setting xlim and ylim in coord_cartesian()."*

*"We can see that reducing the limits is equivalent to subsetting the data. Therefore, to zoom in on a region of the plot, it’s generally best to use coord_cartesian()."*

```{r, include = TRUE}

# Left
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  scale_x_continuous(limits = c(5, 6)) +
  scale_y_continuous(limits = c(10, 25))

# Right
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(color = drv)) +
  geom_smooth() +
  coord_cartesian(xlim = c(5, 6), ylim = c(10, 25))

```

## Themes

There are various themes you can add:

- theme_bw()
- theme_classic(): axes but no grid lines
- theme_dark() 
- theme_gray()
- theme_light()
- theme_linedraw()
- theme_minimal()
- theme_void(): no axes

*"customization of the legend box and plot title elements of the theme are done with element_ functions"*

## Layout

```{r, include = TRUE}

p1 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 1")

p2 <- ggplot(mpg, aes(x = drv, y = hwy)) + 
  geom_boxplot() + 
  labs(title = "Plot 2")

p1 + p2

```

*"It’s important to note that in the above code chunk we did not use a new function from the patchwork package. Instead, the package added a new functionality to the + operator."*

```{r, include = TRUE}

p3 <- ggplot(mpg, aes(x = cty, y = hwy)) + 
  geom_point() + 
  labs(title = "Plot 3")

(p1 | p3) / p2

```

```{r, include = TRUE}

p1 <- ggplot(mpg, aes(x = drv, y = cty, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 1")

p2 <- ggplot(mpg, aes(x = drv, y = hwy, color = drv)) + 
  geom_boxplot(show.legend = FALSE) + 
  labs(title = "Plot 2")

p3 <- ggplot(mpg, aes(x = cty, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 3")

p4 <- ggplot(mpg, aes(x = hwy, color = drv, fill = drv)) + 
  geom_density(alpha = 0.5) + 
  labs(title = "Plot 4")

p5 <- ggplot(mpg, aes(x = cty, y = hwy, color = drv)) + 
  geom_point(show.legend = FALSE) + 
  facet_wrap(~drv) +
  labs(title = "Plot 5")

(guide_area() / (p1 + p2) / (p3 + p4) / p5) +
  plot_annotation(
    title = "City and highway mileage for cars with different drive trains",
    caption = "Source: https://fueleconomy.gov."
  ) +
  plot_layout(
    guides = "collect",
    heights = c(1, 3, 2, 4)
    ) &
  theme(legend.position = "top")

```
