---
title: "Strings"
author: "Joe Shaw"
date: "27/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(babynames)

```

All stringr functions start with str_

*"Beware that the printed representation of a string is not the same as the string itself because the printed representation shows the escapes (in other words, when you print a string, you can copy and paste the output to recreate that string)."*

```{r, include = TRUE}

# String itself - contains two backslashes
string <- "abcdef\\ghi"

# Printed representation of a string (which they refer to as the "raw contents") which has only one backslash
str_view(string)

```

```{r, include = TRUE}

tricky <- "double_quote <- \"\\\"\" # or '\"'
single_quote <- '\\'' # or \"'\""
str_view(tricky)

# Add a "raw string". You put r"()" around your string, or you can use r"{}" or r"[]".
# This means you don't have to keep adding in backslashes which quickly make things unreadable.
less_tricky <- r"(double_quote <- "\"" # or '"'
single_quote <- '\'' # or "'")"
str_view(less_tricky)

# The printed representation is the same, but the raw string version is easier on your eyes when looking at the original string.
str_view(tricky) == str_view(less_tricky)

```
*" You’ll also sometimes see strings containing Unicode escapes that start with \u or \U. This is a way of writing non-English characters that work on all systems."*

```{r, include = TRUE}

x <- c("one\ntwo", "one\ttwo", "\u00b5", "\U0001f604")

x

str_view(x)

```

## 15.2.4 Exercises

```{r, include = TRUE}

# Create strings that contain the following values:

# He said "That's amazing!"
x <- r"(He said "That's amazing!")"

str_view(x)

# \a\b\c\d
y <- r"(\a\b\c\d)"

str_view(y)

#\\\\\\
# 6 backslashes
# Using raw string
z <- r"(\\\\\\)"

str_view(z)

# Wihtout using raw string - you need 12 backslashes, which creates "leaning toothpick sydrome"
z <- "\\\\\\\\\\\\"

str_view(z)

x <- "This\u00a0is\u00a0tricky"

str_view(x)


```

## Creating many strings from data

Concatenating strings with str_c, which is very similar to base::paste0

```{r, include = TRUE}

str_c("x", "y")

str_c("x", "y", "z")

str_c("Hello ", c("John", "Susan"))

?sub()

```

str_flatten works in a similar way to base::paste with collapse 

## Splitting long strings

Useful functions for splitting long strings into rows or columns are:

separate_longer_delim()

separate_longer_position()

separate_wider_delim()

separate_wider_position()

The rest of the chapter is useful for reference but not directly applicable to my work at the moment.
