---
title: "Dates and Times"
author: "Joe Shaw"
date: "27/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(nycflights13)

```

## Dates

"The basic POSIX measure of time, calendar time, is the number of seconds since the beginning of 1970"

POSIX: Portable Operating System Interface. A family of standards to maintain compatibility between operating systems.

POSIXct - the time class in R for calendar time

POSIXlt - the time class in R for local time

POSIXt - a "super class" of ct and lt together
 
*"Thus from 1.3.0, calendar-time objects have class c("POSIXt", "POSIXct"), and broken-down-time objects have class c("POSIXt", "POSIXlt")."*

https://www.r-project.org/doc/Rnews/Rnews_2001-2.pdf 

## Dates and time formats

*"There are three types of date/time data that refer to an instant in time:"*

*"A date. Tibbles print this as <date>."*

*"A time within a day. Tibbles print this as <time>."*

*"A date-time is a date plus a time: it uniquely identifies an instant in time (typically to the nearest second). Tibbles print this as <dttm>. Base R calls these POSIXct, but doesn’t exactly trip off the tongue."*

*"You should always use the simplest possible data type that works for your needs. That means if you can use a date instead of a date-time, you should. Date-times are substantially more complicated because of the need to handle time zones"*

Lubridate has helper functions which convert strings into dates after you provide what order they are formatted in.

```{r, include = TRUE}

ymd("2017-01-31")

mdy("January 31st, 2017")

dmy("31-Jan-2017")


```

UTC = GMT

*"You might wonder what UTC stands for. It’s a compromise between the English “Coordinated Universal Time” and French “Temps Universel Coordonné”"*

make_date() or make_datetime() can be used to create single date or datetime variables from dates that are separated into multiple columns.

You can use modular arithmetic to get hours and minutes out of dates in weird formats.

```{r, include = TRUE}

make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights |> 
  filter(!is.na(dep_time), !is.na(arr_time)) |> 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) |> 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt

```

There are several accessor functions that let you pull individual components out of dates:

-year()
-month()
-mday(): day of the month
-yday(): day of the year
-wday(): day of the week
-hour()
-minute()
-second()

You can also use floor_date and ceiling_date to round date values.

Lubridate includes the option to calculate the difference between times as **durations**, which are defined in seconds.

Alternatively, you can use **periods** which are defined in days and months.

An **interval** is a pair of start and end date-times.

## Time Zones

*"You might wonder why the time zone uses a city, when typically you think of time zones as associated with a country or region within a country. This is because the IANA database has to record decades worth of time zone rules. Over the course of decades, countries change names (or break apart) fairly frequently, but city names tend to stay the same."*

