---
title: "Missing Values"
author: "Joe Shaw"
date: "15/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)

```

## Explicit Missing Values

tidyr::fill() can be used to fill in the gaps for data which has been manually collated. People often leave values bank if they are the same as in the previous row.

```{r, include = TRUE}

treatment <- tribble(
  ~person,           ~treatment, ~response,
  "Derrick Whitmore", 1,         7,
  NA,                 2,         10,
  NA,                 3,         NA,
  "Katherine Burke",  1,         4
)

treatment |>
  fill(everything())


```

You can use dplyr::coalesce() to replace NA with a known value.

```{r, include = TRUE}

x <- c(1, 3, 4, 5, NA, 35, NA, 1)

coalesce(x, 0)

```

## NaN

NaN: not a number

Behaves like NA most of the time.

*"You’ll generally encounter a NaN when you perform a mathematical operation that has an indeterminate result"*

## Implicit Missing Values

*"An explicit missing value is the presence of an absence."*

*"An implicit missing value is the absence of a presence."*

One way to make implicit missing values explicit is to using pivoting.

Another option is to use tidyr::complete()

Another way is to used anti-joins using dplyr::anti_join(x, y)

## Factors

Sometimes factor groups will dissapear if there aren't any observations of that group.

To reveal this, set drop = FALSE in count.

```{r, include = TRUE}

health <- tibble(
  name   = c("Ikaia", "Oletta", "Leriah", "Dashay", "Tresaun"),
  smoker = factor(c("no", "no", "no", "no", "no"), levels = c("yes", "no")),
  age    = c(34, 88, 75, 47, 56),
)

way1 <- health |>  count(smoker)

way2 <- health |>  count(smoker, .drop = FALSE)

way1

way2

```