---
title: "Joins"
author: "Joe Shaw"
date: "30/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(nycflights13)

```

## Definitions

*"**Mutating joins**, which add new variables to one data frame from matching observations in another."*

*"**Filtering joins**, which filter observations from one data frame based on whether or not they match an observation in another."*

**Natural join**: when all variables that appear in both data frames are used as the join key. This occurs when you don't specify the key with the join_by argument.

**Equi join**: when the two foreign keys have the same name.

**Semi join**: a type of filtering join which keeps all the rows in x that have a match in y.

**Anti join**: the opposite of a semi join - all the rows in x that do not have a match in y are kept.

**Key**: the variable used to connect a pair of dataframes together in a join.

**Primary key**: a variable or set of variables that uniquely identifies each observation (each row if the data is tidy) in the original table.

**Foreign key**: a variable or set of variables that corresponds to a primary key in another table.

**Compound key**: when there is more than one variable in a key.

**Surrogate key**: a new variable added to act as a primary key (such as row number).

Note: foreign keys may not always have the same name as the corresponding primary key.

A good way to check primary keys is to use count and make sure that n is never above 1 (i.e. the variable identifies rows uniquely).

You should also check that they aren't any NA values in your primary key.

## Basic Joins

The six basic dplyr joins are:

1. left_join

2. inner_join

3. right_join

4. full_join

5. semi_join

6. anti_join

*"A mutating join allows you to combine variables from two data frames: it first matches observations by their keys, then copies across variables from one data frame to the other."*

*"The primary use of left_join() is to add in additional metadata."*

```{r, include = TRUE}

flights2 <- flights |> 
  select(year, time_hour, origin, dest, tailnum, carrier)

join_by_dest <- flights2 |> 
  left_join(airports, join_by(dest == faa))

join_by_origin <- flights2 |> 
  left_join(airports, join_by(origin == faa))

# The two joins produce different results
all.equal(join_by_dest, join_by_origin) == TRUE


```

## Filtering Joins

Anti-joins are useful for finding *implicit* missing values in datasets. An implicit missing value is one that is not in the dataset but does not have an NA value - it only exists as an absence, not as the presence of an NA.

```{r, include = TRUE}

# Which destinations in the flights2 dataset don't appear in the airports dataset?
flights2 |> 
  anti_join(airports, join_by(dest == faa)) |> 
  distinct(dest)


```

## 20.3.4 Exercises

```{r, include = TRUE}

# Find the 48 hours (over the course of the whole year) that have the worst delays. Cross-reference it with the # weather data. Can you see any patterns?


```

## How Joins Work

*"There is, however, one particularly dangerous case which can cause a combinatorial explosion of rows."*

Rows in x match multiple rows in y.

This is the "many-to-many" join, which dplyr will flag as a warning.

*"Filtering joins never duplicate rows like mutating joins do."*

## Non Equi Joins

*"When we move away from equi joins we’ll always show the keys, because the key values will often be different."*

*"dplyr’s join functions understand this distinction equi and non-equi joins so will always show both keys when you perform a non-equi join."*

There are 4 types of non equi joins.

*"Cross joins match every pair of rows."*

*"Inequality joins use <, <=, >, and >= instead of =="*

*"Rolling joins are similar to inequality joins but only find the closest match."*

*"Overlap joins are a special type of inequality join designed to work with ranges."*

Example with overlap joins: you have a series of parties in an office, and the birthdays of every employee. An overlap join allows you to work out which party should be used to celebrate an employee's birthday: you use a inner_join and the "between" helper.

```{r, include = FALSE}

employees <- tibble(
  name = sample(babynames::babynames$name, 100),
  birthday = ymd("2022-01-01") + (sample(365, 100, replace = TRUE) - 1)
)

parties <- tibble(
  q = 1:4,
  party = ymd(c("2022-01-10", "2022-04-04", "2022-07-11", "2022-10-03")),
  start = ymd(c("2022-01-01", "2022-04-04", "2022-07-11", "2022-10-03")),
  end = ymd(c("2022-04-03", "2022-07-10", "2022-10-02", "2022-12-31"))
)

employees |> 
  inner_join(parties, join_by(between(birthday, start, end)), unmatched = "error")

```