---
title: "Functions"
author: "Joe Shaw"
date: "30/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(nycflights13)

```

*"A good rule of thumb is to consider writing a function whenever you’ve copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)."*

*"To write a function you need to first analyse your repeated code to figure what parts are constant and what parts vary."*

## The Indirection Problem

The indirection problem is easiest to understand with an example.

```{r, include = TRUE}

grouped_mean <- function(df, group_var, mean_var) {
  df |> 
    group_by(group_var) |> 
    summarize(mean(mean_var))
}

# This doesn't work and throws an error
# grouped_mean(diamonds, cut, carat)

```

The error happens because the function looks for "group_var" as a column in the diamonds dataframe, when really it was meant to look for the "cut" column which was supplied to the "group_var" argument.

I've come across this a few times in ddpcr_nipd.

The reason that the indirection problem occurs is due to "tidy evaluation": you can refer to the names of values within dataframes without special treatment.

*"Tidy evaluation is great 95% of the time because it makes your data analyses very concise as you never have to say which data frame a variable comes from; it’s obvious from the context. The downside of tidy evaluation comes when we want to wrap up repeated tidyverse code into a function. "*

*"Tidy evaluation includes a solution to this problem called embracing"*

*"Embracing a variable tells dplyr to use the value stored inside the argument, not the argument as the literal variable name."*

*"{{ var }} will make a dplyr function look inside of var rather than looking for a variable called var."*

```{r, include = TRUE}

grouped_mean <- function(df, group_var, mean_var) {
  df |> 
    group_by({{ group_var }}) |> 
    summarize(mean({{ mean_var }}))
}

grouped_mean(diamonds, cut, carat)

```

You can "wrap" functions inside other functions. A function which uses summarise can be have a grouped dataframe piped into it like it was summarise, and have it give a grouped outcome.

## Data Masking

https://rlang.r-lib.org/reference/topic-data-mask.html

```{r, include = TRUE}

# Defining an environment variable
cyl <- 1000

# Specifying a dataframe column
mean(mtcars$cyl)

# Referring to cyl within a function call causes R to pick the cyl column instead of the cyl value of 1000. The option in the "data" is chosen first and "masks" the option in the environment - "data masking".
dplyr::summarise(mtcars, mean(cyl))

```

## Plot functions

*"We have to use a new operator here, :=, because we are generating the variable name based on user-supplied data. Variable names go on the left hand side of =, but R’s syntax doesn’t allow anything to the left of = except for a single literal name. To work around this problem, we use the special operator := which tidy evaluation treats in exactly the same way as =."*

*"rlang is a low-level package that’s used by just about every other package in the tidyverse because it implements tidy evaluation (as well as many other useful tools)."*

You can have a function which allows the user to supply a condition as an argument.

```{r, include = TRUE}

conditional_bars <- function(df, condition, var) {
  df |> 
    filter({{ condition }}) |> 
    ggplot(aes(x = {{ var }})) + 
    geom_bar()
}

diamonds |> conditional_bars(cut == "Good", clarity)

```

## Labelling

*"rlang is a low-level package that’s used by just about every other package in the tidyverse because it implements tidy evaluation (as well as many other useful tools)."*

*"To solve the labeling problem we can use rlang::englue(). This works similarly to str_glue(), so any value wrapped in { } will be inserted into the string. But it also understands {{ }}, which automatically inserts the appropriate variable name"*

```{r, include = TRUE}

histogram <- function(df, var, binwidth) {
  label <- rlang::englue("A histogram of {{var}} with binwidth {binwidth}")
  
  df |> 
    ggplot(aes(x = {{ var }})) + 
    geom_histogram(binwidth = binwidth) + 
    labs(title = label)
}

diamonds |> histogram(carat, 0.1)

```

## Style

*"R doesn’t care what your function or arguments are called but the names make a big difference for humans. Ideally, the name of your function will be short, but clearly evoke what the function does."*
