---
title: "Base R"
author: "Joe Shaw"
date: "15/11/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

*"This is a good place to remind you that the tidyverse is not the only way to solve data science problems. We teach the tidyverse in this book because tidyverse packages share a common design philosophy, increasing the consistency across functions, and making each new function or package a little easier to learn and use. It’s not possible to use the tidyverse without using base R"*

*"It’s 100% okay to write code that uses a mix of approaches, and don’t let anyone tell you otherwise!"*

## Subsetting with []

df[rows, cols]

*"There’s an important difference between tibbles and data frames when it comes to [. In this book, we’ve mainly used tibbles, which are data frames, but they tweak some behaviors to make your life a little easier. In most places, you can use “tibble” and “data frame” interchangeably, so when we want to draw particular attention to R’s built-in data frame, we’ll write data.frame. If df is a data.frame, then df[, cols] will return a vector if col selects a single column and a data frame if it selects more than one column. If df is a tibble, then [ will always return a tibble."*

Filter is the equivalent of subsetting using square brackets.

```{r, include = TRUE}

df <- tibble(
  x = c(2, 3, 1, 1, NA), 
  y = letters[1:5], 
  z = runif(5)
)

df |> filter(x > 1)

df[!is.na(df$x) & df$x > 1, ]

```

## Subsetting with [[]]

*"[[ and $ can be used to extract columns out of a data frame. [[ can access by position or by name, and $ is specialized for access by name"*

*"There are a couple of important differences between tibbles and base data.frames when it comes to $. Data frames match the prefix of any variable names (so-called partial matching) and don’t complain if a column doesn’t exist"*

*"Tibbles are more strict: they only ever match variable names exactly and they will generate a warning if the column you are trying to access doesn’t exist"*

## Lists

*"The difference between [ and [[ is particularly important for lists because [[ drills down into the list while [ returns a new, smaller list."*

*"This same principle applies when you use 1d [ with a data frame: df["x"] returns a one-column data frame and df[["x"]] returns a vector."*

## Apply

*"The most important member of this family is lapply(), which is very similar to purrr::map()"*

## for loops

*"Rather than making a list and saving the results as we go, a simpler approach is to build up the data frame piece-by-piece:"*

```{r, include = TRUE}

out <- NULL

grow_results <- function() {
  
  for (path in paths) {
    out <- rbind(out, readxl::read_excel(path))
  }
  
}

```

*"We recommend avoiding this pattern because it can become very slow when the vector is very long. This is the source of the persistent canard that for loops are slow: they’re not, but iteratively growing a vector is."*

Base R functions *"tend to have more of a “vector” flavor than a “data frame” flavor because base R functions tend to take individual vectors, rather than a data frame and some column specification."*
