---
title: "Programming as a way of thinking"
author: "Joe Shaw"
format: html
---


"This way of thinking combines some of the best features of mathematics, engineering, and natural science. Like mathematicians, computer scientists use formal languages to denote ideas – specifically computations. Like engineers, they design things, assembling components into systems and evaluating trade-offs among alternatives. Like scientists, they observe the behavior of complex systems, form hypotheses, and test predictions."

"Natural languages are the languages people speak, like English, Spanish, and French. They were not designed by people; they evolved naturally."

"Formal languages are languages that are designed by people for specific applications."

# Maths in Python and R

You can do maths in Python, just like in R. You just need to be aware of when the languages are similar, and when they differ.

## Exponents

```{python}
#| label: python-exponents
#| include: TRUE

# ** is exponent
6 ** 2

```

```{r}
#| label: r-exponents
#| include: TRUE

# ^ is exponent
6 ^ 2

```

## Rounding

```{python}
#| label: python-rounding
#| include: TRUE

round(42.634, ndigits = 1)

```

```{r}
#| label: r-rounding
#| include: TRUE

round(42.634, digits = 1)

```

## Getting help

Finding help documentation is the same.

```{python}
#| label: python-help
#| include: TRUE

help(round)

```

```{r}
#| label: r-help
#| include: TRUE

help(round)

?round

```

## Absolute numbers

```{python}
#| label: python-abs
#| include: TRUE

abs(-42)

```

```{r}
#| label: r-abs
#| include: TRUE

abs(-42)

```

## Strings

In Python and R you can use single ' or double " quotation marks.

```{python}
#| label: python-strings
#| include: TRUE

"This is Python" + " code"

'This is still' + ' python code'

len("Joseph")

```

```{r}
#| label: r-strings
#| include: TRUE

paste0("And this is R", " code")

paste0('And this is still R', ' code')

# length in R returns the length of the vector, not the string
length("Joseph")

stringr::str_length("Joseph")

```

## Types

Python has floating-point numbers, whereas R refers to integers and floats as "doubles".

```{python}
#| label: python-types
#| include: TRUE

type(2)

type(2.0)

```

```{r}
#| label: r-types
#| include: TRUE

typeof(2)

typeof(2.0)

```