---
title: "Quarto"
author: "Joe Shaw"
format:
  pdf: default
editor: visual
---

```{r}
#| include: FALSE
#| warning: FALSE

library(tidyverse)

```

# Notes from [Quarto documentation](https://quarto.org/docs/)

You can include links without square brackets.
Example <https://quarto.org>.

You can embed videos.

# Code blocks, divs and spans

A code block is marked with \`\`\`

```{r}
# include: TRUE

apple <- 1

1 + apple

```

The three-colon ::: symbol marks a fenced division (fenced div).

::: border
This is a div styled with a border
:::

Square brackets are used to define a span (which is a term from HMTL).

[This is a span containing]{.class key="val"}*some italic text.*

Divs are special entities which allow you to control the formatting of a **region** of a document.
Spans are special entities which allow you to control the formatting of a section of **text**.

# Figures

```{r}
#| echo: false
#| label: fig-plot
#| cap-location: top
#| fig-cap: "This is my caption at the top of my figure"

ggplot(mtcars, aes(mpg, cyl)) +
  geom_point() +
  labs(title = "This is my figure title in R")

```

# Using RStudio visual editor

In the visual editor you can highlight text and format it with keyboard shortcuts, and then the source code will automatically update.

**bold using control B**

*italics using control I*

[Link](href) using control K

You can also cross reference like this (@fig-plot) using the visual editor.

The visual editor is a GUI which feels more like Word - it means you don't have to constantly write MarkDown code - the editor translates what you want things to look like back into MarkDown automatically.
This is known as "What You See Is What You Get" (WYSIWYG).

```{r}
# Here is an R code block generated using the editor
```

There is a spell checker in the visual editor.

You can add comments in the visual editor as you would in Word.
This could be a way to get non-R coders to review Quarto documents.

You can even add emojis which is nice 😄

A really useful shortcut is using the forward slash to insert anything.
